Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-02-23 chemotaxis/paper/draft1.tex
made small edits to chemotaxis paper

We consider a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$, and we are interested in uncovering a parameterization of this manifold. 
%
Many algorithms have been proposed for such a task.
%
Historically, principal component analysis (PCA) \cite{shlens2005tutorial} has been used to parmeterize linear manifolds (planes and hyperplanes), where the principal components parameterize the underlying hyperplane.
%
However, in the case when the manifold is curved, such linear techniques can fail to capture the underlying structure.

In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator on the manifold with Neumann boundary conditions \cite{....} can provide coordinates which parameterize the manifold. 
%
To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_x$ and $L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{strip_cnts1}
\includegraphics[width=0.5\textwidth]{strip_cnts2}
\end{figure}

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are often given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions. 
%
The eigenvectors of this matrix then approximate the eigenfunctions of this operator.

Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathbb{R}^n$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$, such that
$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
%
The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues of $A$, $\mu_0, \mu_1, \dots, \mu_{m-1}$, by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

However, the data need not be sampeld uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.

\begin{figure}
\includegraphics[width=0.5\textwidth]{strip_discrete1}
\includegraphics[width=0.5\textwidth]{strip_discrete2}
\includegraphics[width=0.5\textwidth]{strip_nonuniform1}
\includegraphics[width=0.5\textwidth]{strip_nonuniform2}

\end{figure}

@@

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
%
Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $|mathcal{M}_d$. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
\end{equation}
%
and take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
%
The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.

@@

The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
%
To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
%
%In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
%
Automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
%
In this way, we can deduce how many variables are required to write an accurate reduced model.

Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
%
For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
%
However, in general, this mapping from microscale to macroscale is not always obvious.
%
We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
%
Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.

@@

We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles.

@@

Appropriate observers can be constructed in several different ways. 
%
For simple data, such as the illustrative examples in the previous section, the observers are simply the data points themselves. 
%
However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
%
For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
%
For example, when looking at a protein, the ``observers'' may be the locations of the $\alpha$-carbon atoms. 
%
Observers can also factor out other sources of variation. 
%
For example, the bispectrum \cite{...} and the scattering transform \cite{...} are both observers which are invariant to rotations. 
%
Observers should also be robust to noise. 
%
The scattering transform \cite{...} is robust to small deformations, and histograms serve to linearize noise \cite{...}.

After constructing appropriate observers, we must consider what metric to use in the diffusion maps calculation. 
%
Again, the simplest option is to use the Euclidean distance, but this can often prove uninformative. 
%
For molecular simulation data, it is often appropriate to use the root-mean-squared displacement {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
%
For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{...}.
%
Another option is the Earth Mover's Distance, a multi scale distance metric which is stable and robust to noise.

@@

@@

We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
%
We can also use these eigenvalues to detect changes in dynamical behavior.
%
We showed that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there is a gap between the two eigenvalues, as one of the two macroscopic variables is significantly more important to the dynamical behavior.
%
For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.

@@

\subsection{Detect dimensionality}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
We plot  
\begin{equation}
 \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
\end{equation}
when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
%
This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.

>>>>

We first consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes,  the principal axes parameterize the manifold.
%
However, in the case when the manifold is curved, it is more difficult to define such a set of coordinates.

In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions on the manifold \cite{....} can parameterize the manifold. 
%
To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_x$ and $L_y$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_cnts2}
\end{subfigure}
\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ .}
\label{fig:strip_efuncs}
\end{figure}

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions. 
%
The eigenvectors of this matrix then approximate the eigenfunctions of this operator.

Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathcal{M}_d$. 
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator (with Neumann boundary conditions) on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}
%
where $t$ is ....
%
As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

However, the data need not be sampled uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density (TODO: check density vs. potential) $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$. 
%
If the sampling density of the data on the manifold is not meaningful, it can also be factored out via the appropriate normalization of the matrix $W$. TODO: add details?

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_discrete2}
\end{subfigure}

\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\end{subfigure}
\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps. (c, d)Two-dimensional strip with data sampled from a Gaussian distribution in the x-direction and uniform sampling in the y-direction, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps. Note that in both cases we uncover parameterizations which are one-to-one with the x- and y-directions. }
\label{fig:strip_evecs}
\end{figure}

@@

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the leading eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set.
%
Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics will yield the most parsimonious representation of the data.
%
Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.

@@

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
\end{equation}
%
and take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
%
The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...} (TODO: add details in appendix/at end?).

@@

We will now demonstrate our methodology on simulation data from a model problem arising in the study of cellular dynamics. 
%
Although the system consists of many stochastic and noisy particles on the microscale, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
%
However, in general, this mapping from microscale to macroscale is not always obvious.
%
We will show how we can use our methodology to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
%
Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.

%For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
%

@@

Our model problem arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
%
Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
%
We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
%
In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
%
At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
%
%We will analyze the dynamics of collections of such cells/particles.

@@

Appropriate observers can be constructed in several different ways. 
%
For simple data, such as the illustrative examples in the previous section, the observers are simply the data points themselves. 
%
However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
%
For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
%
For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbon atoms, as these backbone atoms are known to describe most of the relevant dynamics within the protein. 
%
Observers can also remove unimportant sources of variation; 
for example, the bispectrum \cite{...} and the scattering transform \cite{...} are both invariant to rotations. 
%
Observers should also be robust to noise. 
%
The scattering transform \cite{...} is robust to small deformations, and histograms serve to linearize noise \cite{...}.

After constructing appropriate observers, we must consider what metric to use in the diffusion maps calculation. 
%
Again, the simplest option is to use the Euclidean distance, but this can often prove uninformative. 
%
For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
%
For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{...}.
%
Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise.

@@

\subsubsection{Using the appropriate metric}

@@

\subsubsection{Detecting changes in dynamical behavior}

We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
%
We can also use these eigenvalues to detect changes in dynamical behavior.
%
We expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
%
For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.

@@

\subsection{Detecting changes in dimensionality}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
From the relationship in \ref{...}, we plot  
\begin{equation}
 \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
\end{equation}
when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
%
This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.
